# Plotting "Spatial" Data with ggplot

**Note** that if you, the student, wish to run all the code
yourself, you should download the `inputs` directory as a zipped file by going [here](https://github.com/eriqande/rep-res-eeb-2017/blob/master/inputs.zip) with a web browser and
then clicking the big "Download" button on the right.  Once you have downloaded that, 
unzip it and put the whole `inputs` directory in the current working directory where you are
working with R. 

Today we are going to take a very cursory look at R's facilities for handling _spatial data_
and ways that it can be plotted using ggplot.  When we say "spatial data" we mean data that
are associated with a geographical location on the earth, and typically 
tied to a specific coordinate system that describes their location.  Why is this important?
It turns out that almost all of the geographical information that you can come by on the
Internet is of this type, and furthermore, it is usually stored in a fairly specialized
form that will look like Greek to the newcomer.  Fortunately, R has a few specialized packages
that make it easy to interact with this sort of data, to read it in and write it out, and 
perform spatial operations.  

For years, these types of tasks were typically carried out with proprietary software like
Esri's [ArcGIS](https://www.arcgis.com/features/index.html) package, where, if you go to their
website they will tell you that "making and sharing beautiful maps" is "possible _only_ with
ArcGIS online."  However, you _can_ make and share beautiful maps from similar data sets using R,
which carries the advantage of being reproducible, open-source, free (as in speech), and free 
(as in beer).  It also, is not too terribly hard to learn to deal with spatial data in R, and it
seems to be getting easier all the time.  So, if you already are using R for data analysis, 
the R route for spatial data certainly
seems to be a better approach than paying Esri 500 clams a year for an ArcGIS subscription.

In this session we will cover only simple topics (keep in mind that your instructor is a 
statistical geneticist that only plays with making maps in R on the side), and we will focus
almost exclusively on plotting data rather than on spatial operations (like testing if points 
are in polygons, or clipping, etc).  The topics will be:

1. Vector data types
1. Coordinate reference systems
1. Plotting vector data types using `ggplot`
1. Raster data types

You will need a few packages to work with these data.  If you don't already have 
them, you will want to get a few from CRAN:
```{r eval=FALSE}
install.packages(c("sp", "rgdal", "raster"))
```
and, though `ggspatial` is available on CRAN, I recommend you get the development version from GitHub:
```{r, eval=FALSE}
devtools::install_github("paleolimbot/ggspatial")
```

When learning the functions in this spatial-data ecosystem, it is often difficult to keep straight
which functions come from which packages, so I will often use namespace addressing (i.e., I will
write `rgdal::readOGR` instead of just `readOGR`, or `ggspatial::geom_spatial` instead of simply
`geom_spatial`, etc.) to make that explicit here.

As always, let's load the tidyverse
```{r}
library(tidyverse)
library(rgdal)
library(sp)
library(ggspatial)
```

## Point and Vector Data Types

Geospatial _point_ data is fairly straightforward---each datum can be represented as a point
on the earth.  There are two other main types of what are called "vector" geospatial data.  These
are _spatial lines_ and _spatial polygons_.  Spatial lines are just a series of points which,
when connected together, form a line.  Spatial polygons are just a series of points which, when
connected together (and with the last point getting connected back to the first one) subtend 
an area.  In addition to being tied to locations on the earth, each point, or line, or polygon
is typically also associated with some other data.  For example,
a polygon could be the boundary of a county (like we saw in the previous lecture) in which case 
it could be associated with a name, an area, and a population size, etc.  A line
might represent a section of a stream, in which case it could be associated with 
the name of the river, the number of fish per mile, and its average flow, etc.

Data of this sort are often stored in what are called _shapefiles_.  As [Wikipedia](https://en.wikipedia.org/wiki/Shapefile) 
states it, "The shapefile format is a popular geospatial vector data format for 
geographic information system (GIS) software. It is developed and regulated by Esri
as a (mostly) open specification for data interoperability among Esri and other
GIS software products. The shapefile format can spatially describe vector features: 
points, lines, and polygons, representing, for example, water wells, rivers, and lakes."

Data of this type can be read into R using `rgdal::readOGR`.  Once it has been read into R,
point, line, or polygon data are stored as objects of class: `SpatialPointsDataFrame`,
`SpatialLinesDataFrame` or `SpatialPolygonDataFrame`, respectively.

### Some example vector data

I have downloaded several small data sets about Santa Cruz that are available,
free and open, from the [county GIS website](http://www.co.santa-cruz.ca.us/Departments/GeographicInformationSystems(GIS).aspx).

I have them in the `inputs` directory of the course webpage repository.  Follow the instructions
at the top of this lecture to get them all.  The data sets that we will play with are:

- Street_Lights --- spatial points giving the location of street lights in the county
- Streams --- spatial lines showing rivers and tributaries
- Watersheds --- spatial polygons showing watershed boundaries

Shapefiles are actually directories that contain a number of different files with different
extensions inside them.
Let's look at the Watersheds shapefile for an example:
```{r}
dir("inputs/santa-cruz-county/Watersheds/")
```
These different files carry all the information.  Some of them are text readable, like the `.prj` 
file, but many of the others are compressed binary files.  Fortunately, we don't ever have to 
open them up and try to read them with a text editor, as we can just read them in with 
`rgdal::readORG`.

### Reading in Vector Spatial Data

The `readOGR` function in `rgdal` does this for us.  It takes as the first argument (the `dsn` parameter) the _directory_ that holds the
shapefile.  In the above case that would be `"inputs/santa-cruz-county/Watersheds"`.  The second argument (the `layer` parameter)
is the prefix of the name of the files that you want to read.  Above, all the files started with `Watersheds` so this second argument
is `Watersheds`.  Let's do it for the street lights, streams, and watersheds:
```{r}
lights <- rgdal::readOGR(dsn = "inputs/santa-cruz-county/Street_Lights", layer = "Street_Lights")
streams <- rgdal::readOGR(dsn = "inputs/santa-cruz-county/Streams", layer = "Streams")
watersheds <- rgdal::readOGR(dsn = "inputs/santa-cruz-county/Watersheds", layer = "Watersheds")
```
When the data are read in, a message about what is in the data set is printed.

Note that the files inside the directory will not always have the same name as the directory. 
For example, inside the "Salmon_Streams" shapefile directory you have:
```{r}
dir("inputs/santa-cruz-county/Salmon_Streams")
```

**Exercise:** read the Salmon_Streams data into a variable called `salmon_streams`.

A side-note here: `readOGR` does not seem to do proper tilde expansion of file names, so trying 
to read something on your hard drive that you might usually do, with a 
tilde giving your home directory, will fail:
```{r, eval=FALSE}
# this fails
stream_fail <- rgdal::readOGR(dsn = "~/Documents/git-repos/rep-res-eeb-2017/inputs/santa-cruz-county/Streams", 
                              layer = "Streams")
```
If you want to do that, you need to wrap the directory path inside the `normalizePath()` function, like
this:
```{r eval=FALSE}
# this doesn't fail on my laptop
stream_no_fail <- rgdal::readOGR(dsn = normalizePath("~/Documents/git-repos/rep-res-eeb-2017/inputs/santa-cruz-county/Streams"), 
                              layer = "Streams")
```


### Quick Visualization of our Vector Data

Before we start drilling down into these data sets to see how `Spatial{Points,Lines,Polygons}DataFrame`s are
structured, let's just plot them quickly.  We will use the `ggspatial` package which extends `ggplot2` to
spatial data with a nice syntax.  Basically, that package provides a new geom called `geom_spatial()` that
deals with plotting spatial points, lines, or polygons.  The same function `geom_spatial` is used to plot
points, lines, or polygons, It "looks at"" the spatial data set it is provided to 
determine whether it should plot points, lines, or polygons.

Here are the watershed boundaries:
```{r, message=FALSE}
ggplot() +
  ggspatial::geom_spatial(data = watersheds, fill = NA, colour = "black") +
  theme_void() +
  coord_map()
```

OK, that is pretty cool.  See that these are, as we expect, polygons.

Let's do the streams:
```{r, message=FALSE}
ggplot() +
  ggspatial::geom_spatial(data = streams, colour = "blue") +
  theme_void() +
  coord_map()
```

Yep, those are lines!

Finally, let's plot the street lights, but we will put them over the top of the watersheds
so that we have a little context for them.
```{r, message=FALSE}
ggplot() +
  ggspatial::geom_spatial(data = watersheds, fill = NA, colour = "black") +
  ggspatial::geom_spatial(data = lights, colour = "red", alpha = 0.4) +
  theme_void() +
  coord_map()
```

OK, those are points. That makes sense.

### The Structure of Spatial*DataFrames

When we write `Spatial*DataFrames` we mean, collecively, `SpatialPointsDataFrames`,
`SpatialLinesDataFrames`, and `SpatialPolygonsDataFrames`.  These are the types of objects
that R uses to store spatial information.  Each of these is what is called an 
S4 object type.  S4 objects are like lists in R, but instead of having elements you access
with `$`, you have information in "slots" that you access with `@`.  (What?!  OK, don't worry too
much, you won't typically ever have to access these slots using `@`, but I do want to use it to
talk about the underlying structure of these beasts.)

First, let's look at the slot names in these three different types of Spatial Data Frames:
```{r}
# spatial points
slotNames(lights)

# spatial lines
slotNames(streams)

# spatial polygons
slotNames(watersheds)
```

Aha! They all have:

- `data` which is a data frame of information about each feature
- `bbox` which is information about the spatial extent of features in the file
- `proj4string` which holds information about the projection (coordinate reference system)

Then, the spatial points have `coords` while the spatial lines have `lines` and the spatial 
polygons have `polygons`.

Let us look more closely at the `data` slot in `streams`.  This is just a data frame, and 
so we can look at it as a tibble:
```{r}
as_tibble(streams@data)
```

And what about the `lines` slot?  Well, that is a spatial data object that holds the 
coordinates that you would connect to make each line.  This is a list of things, in which
each element corresponds to a row in `data`.  

### Subsetting Spatial*DataFrames

If you want to pick out only some of the features in a Spatial*DataFrame, based on 
the values in the `data`, you can use `subset`.  It works a little like `filter` from the 
tidyverse.

For an example, let's see if we can keep only those stream segments that are classified as INTERMITTENT:
```{r}
intermitt <- streams %>%
  subset(., STREAMTYPE == "INTERMITTENT")
```

While there were 1514 features in `streams`
```{r}
length(streams)
```

In `intermitt` there are only 290:
```{r}
length(intermitt)
```

We can plot those to see if the spatial distribution of intermittent streams makes sense to us:
```{r}
ggplot() +
  ggspatial::geom_spatial(data = intermitt, colour = "orange") +
  coord_map()
```

Yep, those are creeks that are mostly up higher in the watersheds.

## Coordinate Reference Systems

I am not going to say too much about this.  Anna Nisi directed me to a fantastic
three-page primer on CRSs in R, and I direct people
[there](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf).

`ggspatial` by default converts everything to lat/lon coordinates, so you can almost get away with
ignoring it, but you should read about it.

To learn about the CRS of a Spatial*DataFrame you can do like this:
```{r}
proj4string(streams)
```
That shows us it is a Lambert Conformal Conic projection ("lcc").

If we wanted to transform the coordinate reference system to something else, we can do 
like this:
```{r}
streams_lat_lon <- spTransform(streams, CRS("+init=epsg:4326"))
```
Check it:
```{r}
proj4string(streams_lat_lon)
```

Yep, that is right.


## Plotting things with ggspatial

For a good primer on `ggspatial` find the README at its [GitHub Page](https://github.com/paleolimbot/ggspatial)

The ggspatial package lets you plot spatial objects using a gpplot-like syntax.  You can map aesthetics to the 
columns in the `data` slot.  For example, let's plot `streams` and color them by STREAMTYPE:
```{r, message=FALSE}
ggplot() +
  ggspatial::geom_spatial(data = streams, mapping = aes(colour = STREAMTYPE)) +
  coord_map()
```

That's cool.

**Exercises**

1. Look at the `data` field of `watersheds`
1. Then plot the watersheds in different colors like this:
```{r, message=FALSE, echo=FALSE}
ggplot() +
  geom_spatial(data = watersheds, aes(fill = NAME), alpha = 0.9) +
  coord_map() +
  theme_void()
```

### Want a background?
How about an open streets background layer?  You can add it with a
`ggspatial::geom_osm()` layer.  Different types of maps are available:
```{r}
# different types of OSM maps available:
rosm::osm.types()
```

So, let's put a hillshade background on:
```{r, cache=TRUE}
ggplot() +
  geom_osm(type = "hillshade") + 
  ggspatial::geom_spatial(data = streams, mapping = aes(colour = STREAMTYPE), alpha = 0.5) +
  coord_map() +
  theme_void()
  
```

Cowabunga!

Using ggspatial with open street maps is a bit nicer than using `ggmap` because 
we are not bound to having square plots and it looks like it figures out the appropriate
zoom for the backgound, etc.  Behold the thin slice of shaded
hillside background!
```{r, cache=TRUE}
ggplot() +
  geom_osm(type = "hillshade") + 
  ggspatial::geom_spatial(data = streams, mapping = aes(colour = STREAMTYPE), alpha = 0.5) +
  coord_map(xlim = c(-122.3, -121.6), ylim = c(37.0, 37.1)) +
  theme_void()
```

It is worth pointing out that you will very often use `coord_map()` when working with `ggspatial`, especially
when using Open Street Maps layers under them.

Check this out, even faceting seems to work with ggspatial.

```{r, cache=TRUE}
ggplot() +
  geom_osm(type = "hillshade") + 
  ggspatial::geom_spatial(data = streams, mapping = aes(colour = STREAMTYPE), alpha = 0.5) +
  coord_map() +
  theme_void() +
  facet_wrap(~ STREAMTYPE)
  
```

These background layers are received as a type of spatial data called a _spatial raster_.
You can think of rasters as digital "images"---basically regular grids of pixels.  
`ggspatial` has exploited one of ggplot's facilities for efficiently plotting images
to be able to efficiently plot rasters in the background.  It also provides a function,
`geom_spraster()` that let's us do the same with spatial rasters from anywhere. This
is a great feature that also let's us plot rasters that have been projected to different
coordinate reference systems.  Before we get into this, let's learn a little about rasters.

## Spatial Raster Data 

R has a fantastic package, called `raster`, written by Robert Hijmans (who was a collaborator with 
Kristen when they were both at Berkeley, check 
[this](http://onlinelibrary.wiley.com/doi/10.1111/j.1365-2699.2006.01517.x/full) out!).  The 
`raster` package provides a nice interface for dealing with spatial raster types and 
doing a variety of operations with them.  

We are going to start with an example: shaded relief of Carmel Bay avaiable through
NOAA's [Digital Elevation Model Global Mosaic (Color Shaded Relief)](http://noaa.maps.arcgis.com/home/item.html?id=feb3c625dc094112bb5281c17679c769).
I have already downloaded it to the `inputs` directory because, to be quite honest, obtaining
it through R was not as straightforward as I would have hoped.  This raster has multiple
layers.  It is a color image stored as a multi-layer (or "band") file.
Accordingly we can use the `brick` function to read it in as a "rasterBrick":
```{r}
carmel_bay <- raster::brick("inputs/carmel_bay_bathy.tif") 
```
Once we have done that we can read about it by printing it:
```{r}
carmel_bay
```
That tells us a lot of useful things, like (from the "dimensions" line) there are 3 layers, each with 3.24 million 
cells, on a grid that is 1800 x 1800 cells.  It also gives us information about
the coordinate reference system (on the "coord. ref." line).

That is all well and good.  Now, let us see what that looks like.  `ggspatial` has the function
`geom_spraster_rgb()` for plotting the entire extent of a three-banded raster, interpreting
the bands as red, green and blue. 
```{r, cache=TRUE, fig.width = 12, fig.height=12}
ggplot() +
  ggspatial::geom_spraster_rgb(carmel_bay) +
  coord_fixed()
```

That is pretty, and could conceivably make a nice background for some of Diana's rockfish
plots.

There is another function in `ggspatial` called `annotation_spraster` that plots a raster,
but does not change the plot
boundaries.  This is very useful if you have a lot of points that you wish to plot, and you want
the plot boundaries to be sized to contain all your points, and you, accordingly, only want that
particular piece of your background raster in the plot.  Let's
see it in action by grabbing Diana's rockfish data, but filtering it only to those
points in the Stillwater Cove area.
```{r}
sebastes_stillwater <- readRDS("inputs/sebastes_locations.rds") %>%
  filter(LATITUDE_M > 36.55, 
         LATITUDE_M < 36.575,
         LONGITUDE_M > -121.975,
         LONGITUDE_M < -121.925)
```

Then plot those.  Here they are by themselves:
```{r}
ggplot() +
  geom_point(data = sebastes_stillwater, mapping = aes(x = LONGITUDE_M, y = LATITUDE_M)) +
  coord_quickmap()
```

And here they are with the raster in the background:
```{r, fig.width=12, fig.height=12, cache=TRUE}
ggplot() +
  ggspatial::annotation_spraster(carmel_bay, interpolate = TRUE) +
  geom_point(data = sebastes_stillwater, mapping = aes(x = LONGITUDE_M, y = LATITUDE_M), 
             colour = "yellow",
             alpha = 0.3) +
  coord_fixed()
```

That is pretty cool.  It might have been nice to have downloaded a higher resolution raster,
which is available, but would have been quite large at the full, zoomed out scale.

One very important thing to note here is that when you are using `ggspatial` you can still plot regular
`ggplot2` geoms on top of it.  We happened to have some points in a tibble (not, a SpatialPointsDataFrame)
with Latitudes and Longitudes, so we just hucked 'em on there using `geom_point`. 

## Wrapping Up

Well, we merely scratched the surface of handling and plotting spatial data in R.  There are a lot of resources out
there:

- [RSpatial Book/Tutorial](http://rspatial.org/index.html) This is an outstanding contribution from Robert Hijmans (it appears).
It looks like it is still in progress, but it is quite complete and lucid. 
- [A CRAN tutorial on spatial data in R](https://cran.r-project.org/doc/contrib/intro-spatial-rl.pdf)
- [Melanie Frazier's kick-ass into to coordinate reference systems](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf)
- [R Spatial Cheatsheet](http://www.maths.lancs.ac.uk/~rowlings/Teaching/UseR2012/cheatsheet.html) loaded with good reminders of how to do things
- [Another Cheat Sheet](http://www.nickeubank.com/wp-content/uploads/2015/10/gis_in_r_vector_cheatsheet.pdf) Great to read through to quickly get a sense of what is possible.
- [ArcGIS_to_R_Spatial_CheatShee](http://www.seascapemodels.org/data/ArcGIS_to_R_Spatial_CheatSheet.pdf) another cheat sheet well-suited to people who have used the ArcGIS point-and-click GUI.
- [Remote sensing in R reference card](https://github.com/wegmann/RSdocs/raw/master/RSecology_cheatsheet/RS_ecology_refcard.pdf) 
Superb summary from the folks who bring you the book [Remote Sensing and GIS for Ecologists Using Open Source 
Software](https://www.amazon.com/Remote-Sensing-GIS-Ecologists-Software/dp/1784270229) which looks like it would be a
superb read if you really wanted to dive deeply into this topic.

On top of that, we haven't even touched on interactive maps with R.  Those interested in that should
check out [Leaflet for R](https://rstudio.github.io/leaflet/).

